The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-07-05 paper/paper.Rnw
Updated README with step-by-step of repo + deleted all useless scripts from before

@@

<<loading_data-recoding>>=
pisa_all <- read_rds("./data/pisa_listcol.Rdata")

@@

% This chunk needs to have cache = TRUE once you want to run it for like 2 hours until 
% the results_* models have been created once and the cache can save them

<<modeling>>=
adapted_year_data <-
    map(pisa_all2$value, ~ {
      if (unique(.x$wave) == "pisa2000") {
        # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
        # is 7 and the bottom is 1
        .x <-
          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
      } else {
        .x <-
          mutate(.x, new_hisced = as.character(high_edu_broad))
      }
      .x
})

@@

% Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
% Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
% Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
% Talk more about how countries with high social mobility has been linked to smallest achievement gaps
% Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.

@@

% Show graph of variance over time
% I'm showing standard deviations, is it better to show variance?

@@

var_years %>%
  setNames(seq(2000, 2015, 3)) %>%
  enframer("name") %>%
  ggplot(aes(name, value)) +
  geom_col() +
  scale_y_continuous(name = "Standard deviation of raw test scores") +
  scale_x_discrete(name = NULL) +
  ggtitle("Standard deviation of test scores over time") +
  theme_few() +
  theme(text = element_text(size = 12))
@
\caption{Standard deviation of test scores across all waves.}
\end{center}
\end{figure}

@@

@@

<<table_escs, include = TRUE>>=

@@

\subsection{Evolution of the achievement gap}
% Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).

@@

<<evolution_reading, eval = F>>=
# See below that reading graph is upon request. This is the one.
complete_data_topbottom %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "read", !is.na(country)) %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  geom_linerange(aes(ymin = 0, ymax = difference))+
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 2), size = 0.7,
            colour = "red") +
  facet_wrap(~ country, ncol = 5) +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  coord_fixed() +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")

@@

% How to output data frames are latex tables?
<<rate_change>>=

@@

<<table_sample>>=
sample_tables_topbottom %>%
  filter(name %in% c(2000, 2015)) %>%
  transmute(name, country, escs_dummy, N = paste0(n, " (", perc, ")")) %>%
  spread(escs_dummy, N) %>%
  rename(Year = name,
         Country = country,
         `10th SES` = `0`,
         `90th SES` = `1`) %>%
  select(Country, Year, everything()) %>%
  arrange(Country, Year) %>%
  xtable::xtable()
@

>>>>

% Things to remember:
% The working directory is not where the data is.
% Change working directory to "/Users/cimentadaj/Downloads/gitrepo/Achievement_Gap_PISA/paper"
% for everything below to work.

% There's a code chunk named modeling which lasts about 3-4 hours to run. I explain how to get
% around that in a comment above the code chunk. For down there.

@@

% pisa_listcol.Rdata is a dataset produced by the source() expression from the above
% code chunk
<<loading_data-recoding>>=
pisa_all <- read_rds("./data/pisa_listcol.Rdata")

@@

% This chunk takes about 3-4 hours to run. Once I had them ready I saved them so you can read them
% as rds. If you run the 'compile pdf' this chunk is already cached (DONT CHANGE ANYTHING FROM
% THIS CHUNK, CAUSE OTHERWISE IT WILL RERUN FOR 3-4 HOURS). If you want to run the code without
% compiling sweave, comment the test_diff expressions and read as rds.

% Remember to uncomment when compiling sweave again, so the chunk remains intact.
<<modeling>>=
adapted_year_data <-
    map(pisa_all2$value, ~ {
      if (unique(.x$wave) == "pisa2000") {
        # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
        # is 7 and the bottom is 1
        .x <-
          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
      } else {
        .x <-
          mutate(.x, new_hisced = as.character(high_edu_broad))
      }
      .x
})

@@

% TODO: Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
% Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
% Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
% Talk more about how countries with high social mobility has been linked to smallest achievement gaps
% Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.

@@

@@

var_years %>%
  setNames(seq(2000, 2015, 3)) %>%
  enframer("name") %>%
  ggplot(aes(name, value)) +
  geom_col() +
  scale_y_continuous(name = "Standard deviation of raw test scores") +
  scale_x_discrete(name = NULL) +
  ggtitle("Standard deviation of test scores over time") +
  theme_few() +
  theme(text = element_text(size = 14))
@
\caption{Standard deviation of test scores across all waves.}
\end{center}
\end{figure}

@@

% Previous table that had % of low and high education for low and high SES dummies

@@

% New table only has sample size, average score and s.e.
<<table_escs, include = TRUE>>=

@@

\subsection{Evolution of the achievement gap}
% TODO: Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).

@@

% Code chunks for the same plots as above for reading. Available upon request.
<<evolution_reading, eval = F>>=
# See below that reading graph is upon request. This is the one.
complete_data_topbottom %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "read", !is.na(country)) %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  geom_linerange(aes(ymin = 0, ymax = difference))+
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 2), size = 0.7,
            colour = "red") +
  facet_wrap(~ country, ncol = 5) +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  coord_fixed() +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")

@@

<<rate_change>>=

@@
